---
title: "Template to create patient profiles for SDTM datasets"
subtitle: "Study: X, Batch X"
author: "Laure Cougnaud"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: glpgStyle::html_report
vignette: >
  %\VignetteIndexEntry{SDTM template for patient profiles}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

This document presents an example of standard patient profile report
for SDTM datasets.

This document is intended to be used as a template to create patient profile
report for your specific study.
You can copy the source code of this document into your working directory with:
```{r setUpTemplateForYourStudy, eval = FALSE}

# get template from the package
pathTemplate <- system.file(
	"doc", "patientProfiles-template-SDTM.Rmd", 
	package = "patientProfilesVis"
)
file.copy(from = pathTemplate, to = ".")
# Note: your current working directory can be checked with: getwd()
```
This template can then be further tailored to your specific study (changing data path,
variables available in the data, ...).

```{r patientProfiles-optionsChunks, echo = FALSE, cache = FALSE}

	library(knitr)

	knitr::opts_chunk$set(
		echo = TRUE, # print code in the report
		message = TRUE,
		error = FALSE, # stop document execution if error (not the default)
		fig.align = "center",
		warning = FALSE,
		# to have list of DT included within a chunk with: glpgUtilityFct::knitPrintListObjects
		results = 'asis'
	)
	
```

The following R packages are required:

```{r patientProfiles-loadPackages}

	library(patientProfilesVis)
	library(glpgUtilityFct)
	library(glpgStyle)

```

## Dataset

The SDTMs datasets are imported into R.

```{r patientProfiles-loadData}
	
	# For this vignette, an example STDM dataset from the Pelican
	# study is imported into R
	data(SDTMDataPelican, package = "glpgUtilityFct")
	data(labelVarsSDTMPelican, package = "glpgUtilityFct")
	dataAll <- SDTMDataPelican
	labelVars <- labelVarsSDTMPelican
	
	# If SDTM datasets should be imported from the external folder,
	# the following code could be used:
#	pathFiles <- list.files(path = "/path/to/data", pattern = "*.sas7bdat$", full.names = TRUE)
#	dataAll <- loadDataADaMSDTM(files = pathFiles)
#	labelVars <- attr(dataAll, "labelVars")

```

## Palettes

Color and shape palette for the reference indicator variable are
created. These shape of the symbols are specified Unicode format in hexadecimal
 (see [List of unicode symbols](https://en.wikipedia.org/wiki/List_of_Unicode_characters)).

```{r patientProfiles-palettes}

	colorPalette <- glpgColor(c("main", "extra", "highlight"))
	
	# color/shape palette for reference range indicator (abnormality lab/vs)
	# Please note that this might be manually updated
	shapePaletteRind <- c(
		`Low Panic` = '\u25BC', # big filled up-pointing arrow
		LOW = '\u25BC', Low = '\u25BC', # filled down-pointing arrow
		`Low Normal` = '\u25BC', # small filled down-pointing arrow
		NORMAL = '\u25CF', Normal = '\u25CF', # filled circle
		`High Normal` = '\u25B2', # small filled up-pointing arrow
		HIGH = '\u25B2', High = '\u25B2', # filled up-pointing arrow
		`High Panic` = '\u25B2', # big filled up-pointing arrow
		Abnormal = '\u2605', # star
		'NA' = '\u271A', # cross (3)
		Unknown = '\u271A'
	)
	colorPaletteRind <- c(
		`Low Panic` = unname(colorPalette["signalRed"]),
		LOW = unname(colorPalette["orange"]), Low = unname(colorPalette["orange"]),
		`Low Normal` = unname(colorPalette["yellow"]),
		NORMAL = unname(colorPalette["green"]), Normal = unname(colorPalette["green"]),
		`High Normal` = unname(colorPalette["yellow"]),
		HIGH = unname(colorPalette["orange"]), High = unname(colorPalette["orange"]),
		`High Panic` = unname(colorPalette["signalRed"]),
		Abnormal = unname(colorPalette["signalRed"]),
		'NA' = unname(colorPalette["grey"]),
		Unknown = unname(colorPalette["grey"])
	)

```

```{r patientProfiles-initialization}

	patientProfilesPlots <- list()

```

# Creation of patient profiles

For each domain of interest, the subject profiles are created.
See the vignette: 

## Demographics

```{r patientProfiles-demographics}

	dmPlots <- subjectProfileTextPlot(
		data = dataAll$DM,
		paramValueVar = c(
			"SEX", "RACE", "COUNTRY", "ARM",
			"SITEID", "AGE", "RFSTDTC"
		),
		# optional:
		labelVars = labelVars
	)
	patientProfilesPlots <- c(patientProfilesPlots, list(DM = dmPlots))

```

## Medical history

```{r patientProfiles-medicalHistory}
	
	mhPlots <- subjectProfileTextPlot(
		data = dataAll$MH,
		paramValueVar = c("MHDECOD", "MHSTDTC", "MHENDTC"),
		# optional:
		paramGroupVar = "MHSTDTC",
		title = "Medical history (Start - End)",
		labelVars = labelVars,
		table = TRUE
	)
	patientProfilesPlots <- c(patientProfilesPlots, list(MH = mhPlots))

```

## Concomitant medications

```{r patientProfiles-concomitantMedications}

	cmPlots <- subjectProfileIntervalPlot(
		data = dataAll$CM,
		paramVar = c("CMDECOD", "CMDOSTXT", "CMDOSU", "CMDOSFRQ", "CMROUTE"),
		timeStartVar = "CMSTDY",
		timeEndVar = "CMENDY",
		# optional:
		paramGroupVar = "CMINDC",
		colorVar = "CMINDC",
		labelVars = labelVars,
		title = "Concomitant medications",
		# To zoom in axis scale in study time frame
		# (to avoid scale is focused on negative pre-study time frame for CM)
		timeTrans = getTimeTrans(type = "asinh-neg"), 
		alpha = 0.8,
		timeAlign = FALSE
	)
	
	patientProfilesPlots <- c(patientProfilesPlots, list(CM = cmPlots))

```


## Treatment exposure

```{r patientProfiles-treatmentExposure}

	exPlots <- subjectProfileIntervalPlot(
		data = dataAll$EX,
		paramVar = c("EXTRT", "EXDOSE", "EXDOSU", "EXDOSFRM", "EXDOSFRQ", "EXROUTE", "EXTPT"),
		timeStartVar = "EXSTDY",
		timeEndVar = "EXENDY",
		# optional:
		colorVar = "EXDOSFRM",
		title = "Treatment exposure",
		timeAlign = FALSE,
		alpha = 0.8,
		labelVars = labelVars
	)

```

## Adverse events

```{r patientProfiles-adverseEvents}

	aePlots <- subjectProfileIntervalPlot(
		data = dataAll$AE,
		paramVar = "AEDECOD",
		timeStartVar = "AESTDY",
		timeEndVar = "AEENDY",
		# optional:
		paramGroupVar = "AESOC",
		colorVar = "AESEV", 
		title = "Adverse events",
		timeAlign = FALSE,
		alpha = 0.8,
		labelVars = labelVars
	)
	
	patientProfilesPlots <- c(patientProfilesPlots, list(AE = aePlots))
	
```

## Laboratory measurements

```{r patientProfiles-laboratory}

	# specify custom color and shape palette
	lbRind <- unique(dataAll$LB$LBNRIND)
	colorPaletteLB <- colorPaletteRind[which(names(colorPaletteRind) %in% c(lbRind, "NA"))]
	shapePaletteLB <- shapePaletteRind[which(names(shapePaletteRind) %in% c(lbRind, "NA"))]
	
	lbPlots <- subjectProfileLinePlot(
		data = dataAll$LB,
		paramValueVar = "LBSTRESN",
		paramNameVar = "LBTEST",
		timeVar = "LBDY",
		# optional
		paramValueRangeVar = c("LBSTNRLO", "LBSTNRHI"),
		paramGroupVar = c("LBCAT", "LBSCAT"),
		colorVar = "LBNRIND", colorPalette = colorPaletteLB,
		shapeVar = "LBNRIND", shapePalette = shapePaletteLB,
		shapeSize = 4,
		title = "Laboratory test measurements",
		alpha = 0.8,
		labelVars = labelVars
	)
	
	patientProfilesPlots <- c(patientProfilesPlots, list(LB = lbPlots))
	
```

## Vital signs

```{r patientProfiles-vitalSigns}

	# If available, reference range indicator
	# could be displayed via the color/shape variable

	vsPlots <- subjectProfileLinePlot(
		data = dataAll$VS,
		paramValueVar = "VSSTRESN",
		paramNameVar = "VSTEST",
		timeVar = "VSDY",
		# optional
		paramGroupVar = "VSCAT",
		colorVar = "VSCAT", 
		shapeSize = 2,
		title = "Vital signs",
		alpha = 0.8,
		labelVars = labelVars
	)
	patientProfilesPlots <- c(patientProfilesPlots, list(VS = vsPlots))
	
```

## Electrocardiogram

```{r patientProfiles-ECG}

	# If available, reference range indicator
	# could be displayed via the color/shape variable

	egPlots <- subjectProfileLinePlot(
		data = dataAll$EG,
		paramValueVar = "EGSTRESN",
		paramNameVar = "EGTEST",
		timeVar = "EGDY",
		# optional
		title = "Electrocardiogram",
		alpha = 0.8,
		labelVars = labelVars
	)
	patientProfilesPlots <- c(patientProfilesPlots, list(EG = egPlots))
	
```

## Creation of the subject profiles

```{r patientProfiles-createSubjectProfileReport, message = FALSE, warning = TRUE}

	pathsPatientProfiles <- createSubjectProfileReport(
			
		listPlots = patientProfilesPlots, 
		
		# optional
		reportPerSubject = TRUE, 
		verbose = TRUE,
		outputFile = './patientProfiles/subjectProfile.pdf',
		timeAlign = "all", timeAlignPerSubject = "all",
		exportBatchSize = 5,
		
		# export subjects with highest AE toxicity grades first
		subjectSortData = dataAll$AE,
		subjectSortVar = "AESEV",
		subjectSortDecreasing = TRUE,
		
#		# only consider subset of patients:
#		# - option 1: from a dataset
#		subjectSubsetData = dataAll$AE,
#		subsetVar = "AETOXGR",
#		subsetValue = c("MODERATE", "SEVERE"),
#		# - or from specified USUBJID
		subset = c("study-4903-07", "study-4908-01", "study-4907-07")

	)

```

# Session information

```{r sessionInformation, echo = FALSE}

	library(pander)
	pander(sessionInfo())

```